Modelling the gas kinematics of an atypical Lyman alpha emitting compact dwarf galaxy

### 2017-03-13 paper/letter.tex
moving parts

\tol\ is a compact star forming dwarf galaxy that presents a
strong \lya\ emission \citep{Thuan97} with two puzzling 
features: the line is symmetric, single peaked and wide around the
line's center.
Usually the \lya\ line has an asymmetric single or double peak. 
These two special features in \tol\ cannot be explained quantitatively
tith conventional shell models
\citep{Mas-Hesse09,2015A&A...578A...7V,2015ApJ...812..123G}.

@@

In this paper we review first the observational characteristics of
\tol, then we summarize the main feature of the multiphase and
rotation models and explain how we fit the free parameters in the
model to the \lya\ line shape in \tol.
We use the kinematic result to interpret them in terms of the
dynamical mass that should be present in \tol.
We close by discussing the physical scenarios that open up by our
results and the additional observational information needed to clarify
the kinematic nature of \tol.

@@

In order to map out this large parameter space, we randomly drew
$2500$ sets of parameters within a observationally realistic range,
based on the considerations of \cite{Laursen2013ApJ...766..124L}, 
yielding a large variety of single-, double- and triple-peaked
spectra. 
The full analysis of the the spectral features as well as
more details on the radiative transfer are presented in
\citep{Gronke2016}.

@@

Then we perform a Kolmogorov-Smirnov test to compare each parameter
distribution in the best $25$ models against the parent distribution
of $2500$ models. 
If we obtain a p-value $<0.05$ for a given model parameter, we conclude that
this parameter does influence the $\chi^2$ fit, as the distribution for
the best $\chi^2$ models is statistically different to the
distribution from the global sample of $2500$ models.

@@

The rotation model corresponds to the work presented in
\citep{GaravitoCamargo2014} based on the Monte Carlo code
\texttt{CLARA} \citep{CLARA}. 
In that model the \lya\ photons are propagated 
within a spherical and homogeneous cloud of HI gas undergoing solid
body rotation.
The sphere is fully characterized by three parameters: the HI line's
center optical  depth $\tau$ measured from the center to its surface, the HI
temperature $T$, and the linear surface velocity $V_{\rm max}$.  
Including the effect of dust only changes the overall line
normalization but not its shape.  
The results we report do not include any dust model.
Here we use an analytical solution that captures the most important
effects of rotation onto the \lya\ line as explained by
\citet{GaravitoCamargo2014} and allows us a full parameter space
exploration.

The analytical solution for the rotation sphere is the base to
perform a Markov Chain Monte Carlo (MCMC) calculation using the
\texttt{emcee} Python library \citep{2013PASP..125..306F}. \texttt{emcee} 
is an open source optimized implementation of the affine-invariant 
ensemble sampler for MCMC. 
The algorithm creates a number of walkers that,
during a sufficient number of steps, generate parameters' combinations
for a specific model.
For each timestep, the code calculates the likelihood of the
combination with respect to the observational data.
The walkers explore
the parameter space sampling the likelihood function built as $\propto
\exp(\chi^2/2)$, where the $\chi^2$ follows the definition in Equation
\ref{eq:chi2}.

@@

Qualitatively, as \tol\ possesses a very wide spectrum which can be
achieved by subsequent scatterings off (relatively) fast moving clumps
while the multi-phase nature (i.e., the existence of low-density
channels) ensures the high flux at line center as observed.
A lower velocity dispersion would produce a thiner line than the
observed.
From the lower right panel in \ref{multiphaseresults} we find that it
is unlikely that the clump velocity dispersion is lower than
$50$\kms.

@@

Using the constrains for the velocity dipersion $\sigma$ and the
maximum rotational velocity $V_m$ in a spherical system localized in a
region of size $r$ we can estimate two different values for the
dynamical mass within $r$:

@@

\tol's stellar mass is  $M_{\star} = 1.45\pm0.45\times 10^{8}
M_{\odot}$   \citep{2014PASP..126.1079M} and its total neutral HI mass is $M_{\rm HI}<2.65\times 10^{8}$ M$_{\odot}$. 
\citep{pustilnikmartin07}; the dynamical mass is at least 12 to 160 times
the baryonic mass, depending if one considers the multiphase or
bulk rotation estimate. 
We lean towards the lower dynamical mass estimate from the multiphase
model as it seems easier to reconcile with the following two
astrophysical mechanisms for its origin.

@@

A new observational test is needed to clarify the kinematical nature of
\tol. 
We suggest that integral field unit measurements 
resolving its spatial extent are up to the task. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer
\citep{2014Msngr.157...13B} with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
We note that archival Chandra X-ray archival
data \footnote{\url{http://cxc.harvard.edu/cda/}} does not show any
detection for \tol.
This lack of detection would also put an upper limit on the intensity
of recent black hole activity due to accretion.

Another observational test includes the measurement of the
\lya\ ionizing continuum escape fraction.
In the rotational model this fraction should be zero, while
the multiphase model predicts that averaging over all sightlines
it should be around $0.5^{+1.0}_{-0.4}$\%, with the possibility of strong
variations depending on viewing angle \citep{Gronke2016}.

@@

In this paper we presented two kinematic models that independently
reproduce the so far unexplained observational features of \tol's
\lya\ emission line.

One model is based on a multiphase ISM with random clump motions and
the other on gas bulk solid body rotation.
It is the first time that an observed \lya\ profile can be fully
reproduced by either of these two kinematic conditions.
Furthermore, the kinematics of both models suggest a dynamical mass
at least ten times larger than the baryonic mass infered from observations.
We suggest that additional integral field unit observations are the
only way to be certain about the detailed kinematic structure in \tol.

All in all, the mere existence of a strong LAE galaxy with a broad,
symmetric line is interesting.
It raises the question whether some high redshift LAEs have asymmetric
lines because the blue half was truncated by the intergalactic medium.
In this case the \lya\ radiation could emerge as a low surface
brightness glow, which may be connected to \lya\ halos, while also
influencing the way LAEs can be used as a probe of reionization
\citep{2014PASA...31...40D}.

These findings demonstrate the importance of including rotation and multiphase
conditions as kinematic features to model the \lya\ line in high redshift
galaxies.
Additionally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it could correspond to a so far undetected
supermassive black hole in a dwarf galaxy, providing a new way to test
and probe theories on the co-evolution of galaxies and black holes in
the first generation of galaxies.

>>>>

\tol\ is a compact star forming dwarf galaxy that presents a
strong \lya\ emission \citep{Thuan97} with two puzzling 
features: the line is symmetric, single peaked and wide around the
line's center.
Usually the \lya\ line has an asymmetric single or double peak. 
These two special features in \tol\ cannot be explained quantitatively
with conventional shell models
\citep{Mas-Hesse09,2015A&A...578A...7V,2015ApJ...812..123G}.

This kind of profile also raises the question whether some high redshift
LAEs have asymmetric lines because the blue half was truncated by the
intergalactic medium \citep{2007MNRAS.377.1175D}. 
In this case the \lya\ radiation could emerge as a low surface
brightness glow, which may be connected to \lya\ halos, while also
influencing the way LAEs can be used as a probe of reionization
\citep{2014PASA...31...40D}.

@@

We review first the observational characteristics of
\tol, then we summarize the main features in the multiphase and
rotation models to explain how we fit their free parameters 
to the \tol's \lya\ line shape.
We use the results to interpret them in terms of the galaxy's
dynamical mass. 
We close by discussing the physical scenarios that open up by our
results and the additional observational information needed to clarify
the kinematic nature of \tol.

@@

In order to map out this large parameter space, we randomly drew
$2500$ sets of parameters within a observationally realistic range,
based on the considerations of \citet{Laursen2013ApJ...766..124L}, 
yielding a large variety of single-, double- and triple-peaked
spectra. 
The full analysis of the the spectral features as well as
more details on the radiative transfer are presented by
\citet{Gronke2016}.

@@

Then we perform a Kolmogorov-Smirnov (K-S) test to compare each parameter
distribution in the best $25$ models against the parent distribution
of $2500$ models. 
If we obtain a p-value $<0.05$ for a given model parameter, we conclude that
this parameter does influence the $\chi^2$ fit, as the distribution for
the best $\chi^2$ models is statistically different to the
distribution from the global sample of $2500$ models.

@@

The rotation model corresponds to the work presented in
\citep{GaravitoCamargo2014} based on the Monte Carlo code
\texttt{CLARA} \citep{CLARA}. 
In that model the \lya\ photons are propagated 
within a spherical and homogeneous cloud of HI gas undergoing solid
body rotation.
The sphere is fully characterized by three parameters: the HI line's
center optical  depth $\tau$ measured from the center to its surface, the HI
temperature $T$, and the linear surface velocity $V_{\rm max}$.  
In this model, dust only changes the overall line
normalization and only weakly its shape.  
The results we report do not include any dust model.
Here we use an analytical solution that captures the most important
effects of rotation onto the \lya\ line as explained by
\citet{GaravitoCamargo2014} and allows us a full parameter space
exploration.

The analytical solution for the rotation sphere is the base to
perform a Markov Chain Monte Carlo (MCMC) calculation using the
\texttt{emcee} Python library \citep{2013PASP..125..306F}. \texttt{emcee} 
is an open source optimized implementation of the affine-invariant 
MCMC sampler \citep{goodman2010ensemble}. 
The algorithm creates a number of walkers that,
during a sufficient number of steps, generate parameters' combinations
for a specific model.
For each timestep, the code calculates the likelihood of the
combination with respect to the observational data.
The walkers explore the parameter space sampling the Gaussian likelihood
function built as $\propto \exp(-\chi^2/2)$, where the $\chi^2$ follows
the definition in Equation \ref{eq:chi2}.

@@

Qualitatively, as \tol\ possesses a very wide spectrum which can be
achieved by subsequent scatterings off (relatively) fast moving clumps
while the multi-phase nature (i.e., the existence of low-density
channels) ensures the high flux at line center as observed.
In turn, a lower velocity dispersion would produce a thiner line than the
observed.
From the lower right panel in \ref{multiphaseresults} we find that it
is unlikely that the clump velocity dispersion is lower than
$50$\kms.

Why is it that the other parameters do not seem to matter? Probably
because $P_{cl}$ (the probability for a \lya\ to be emitted in a
clump) is high and most photons are produced at
large radii making the radiative tranfer effects in the center less
important.  
We cannot discard that another region of parameter spacem with similar
velocity structure but empty interclump medium, relatively dense
clumps and at least a few clumps per sightline 
could also work \citep{Hansen06}.
This possibility does not change our conclusions about the
kinematic parameters, but it might be interesting to explore in the
future with a higher sampling of the parameter space once new
observations clarify the kinematic nature of \tol.

@@

\tol\ shows one of the most atypical \lya\ spectra observed so far.
Its flux at the line's center is high compared to other LAEs at low
redshift and the wide symmetrical peak is virtually absent from other
LAEs at low and high redshift \citep{LARS,Erb14,Trainor16}. Simple
shell models fail to reproduce such a spectrum
\cite{2015A&A...578A...7V}.  
In the previous sections we show how these characteristics can be
explained by rather extreme kinematic characteristics.
Now we discuss and frame those results in terms of the dynamical mass
that could explain the velocities obtained in the model.

Using the constraints for the velocity dipersion $\sigma$ and the
maximum rotational velocity $V_m$ in a spherical system localized in a
region of size $r$ we can estimate two different values for the
dynamical mass within $r$:

@@

\tol's stellar mass is  $M_{\star} = 1.45\pm0.45\times 10^{8}
M_{\odot}$   \citep{2014PASP..126.1079M} and its total neutral HI mass
is $M_{\rm HI}<2.65\times 10^{8}$ M$_{\odot}$ 
\citep{pustilnikmartin07}; the dynamical mass is at least 12 to 160 times
the baryonic mass, depending if one considers the multiphase or
bulk rotation estimate. 
We lean towards the lower dynamical mass estimate from the multiphase
model as it seems easier to reconcile with the following two
astrophysical mechanisms for its origin.

@@

A new observational test is needed to clarify the kinematical nature of
\tol. 
We suggest that integral field unit measurements 
resolving its spatial extent are up to the task. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer (MUSE)
\citep{2014Msngr.157...13B} with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map in the H$\alpha$ line,
following the kind of work presented by \citet{Herenz16} done with the Potsdam
Multi-Aperture Spectrometer (PMAS) \citet{PMAS} on nearby
\lya\ emitting galaxies.
We note that archival Chandra X-ray archival
data \footnote{\url{http://cxc.harvard.edu/cda/}} does not show any
detection for \tol.
This lack of detection would also put an upper limit on the intensity
of recent black hole activity due to accretion.

Another observational test includes the measurement of the
\lya\ ionizing continuum escape fraction.
In the rotational model this fraction should be zero, while
the multiphase model predicts that averaging over all sightlines
it should be around $0.5^{+1.0}_{-0.4}$\%, with the possibility of strong
variations depending on viewing angle (i.e. some sightlines can have
an escape fraction close to $100\%$)\citep{Gronke2016} .

@@

In this paper we presented two kinematic models that independently
reproduce the so far unexplained observational features of \tol's
\lya\ emission line.  
One model is based on a multiphase ISM with random clump motions and
the other on gas bulk solid body rotation.

It is the first time that an observed \lya\ profile can be fully
reproduced by either of these two kinematic conditions.
Furthermore, the kinematics of both models suggest a dynamical mass
at least ten times larger than the baryonic mass infered from observations.
We suggest that additional integral field unit observations are the
only way to be certain about the detailed kinematic structure in \tol.

All in all, the mere existence of a strong LAE galaxy with a broad,
symmetric line is interesting. 
\tol's line shape is different to others and reveals an unexpected
kinematic structure.
Our findings highlight the importance of including multiphase and/or
rotation conditions as kinematic features to model the \lya\ line.

Additionally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it could correspond to a so far undetected
supermassive black hole in a dwarf galaxy, providing a new way to test
and probe theories on the co-evolution of galaxies and black holes in
the first generation of galaxies.
